\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-16 ace-pldi14/ace-pldi14.tex
pldi14 sec 3 draft1 finally done HOLY GOD NO TIME LEFT

Translation of terms of the form \verb|c.ni| and \verb|c.i| operates via \verb|trans_Attribute|. Because we are implementing complex numbers as vector types in OpenCL, we generate an access to the left element, accessed via the attribute \verb|s0|, to access the non-imaginary component and and \verb|s1| to access the imaginary component. The target provides a programmatic API for generating code in the target language via its methods, as can be seen. Translation can assume typechecking succeeded, so the implementation assumes that if the attribute is not \verb|ni|, it is \verb|i|.

It does so by again applying the dispatch protocol to call a method of the form \verb|translate_|$X$, where $X$ is the syntactic form of the expression. This method is responsible for returning a copy of the expression's ast node with an additional attribute, \verb|code|, containing the source code of the translation, represented here as a string though it may also be represented in a structured manner. In our example, it is simply a direct translation to the corresponding OpenCL attribute access (Line 20), using the recursively-determined translations of the operands (Lines 16-17).  More sophisticated abstractions may insert arbitrarily complex statements and expressions during this phase. The context also provides some support for non-local insertions, such as new top-level type declarations, imports and helper code (not shown).

\subsection{Active Targets}\label{atargets}

\subsubsection{Target Types}
Every active type must translate to a unique  \emph{target type} -- a type in the target language (dynamically typed target languages can be thought of as only having a single type, \verb|dyn| \cite{pfpl}). Target types are constructed via methods of a provided {active target}, called \verb|target| throughout this listing. Here, a complex number translates to a pair of values, each having the type that is the target type of \verb|t|, the numeric type that was passed into the constructor (lines \ref{cplx}.7-\ref{cplx}.8). The \verb|tuple| method returns a representation of an actual type (not an active type) in the target language as the return value of the \verb|tuple| method (e.g. 2-tuples of integers compile to the vector type \verb|int2| in OpenCL). This design still allows the complex number type being defined here to supports any target for which tuples can be defined directly however (e.g. CUDA, Python, etc.).

\todo{nathan's work on representational consistency?}

\subsection{Active Bases}\label{abases}
Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface. The active base specifies the \emph{base semantics} of that function -- the semantics of statements and expressions that do not have a clear ``primary'' subexpression.

For example, the type and translation of a numeric literal is determined by the active base. The active base \verb|clx.base|, for example, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, to be consistent with the semantics of OpenCL and C99. The \verb|clx.base| object is an instance of \verb|clx.Base|. This class is written so that this choice can be made differently if desired. For example, the base \verb|clx.Base(flit_t=clx.float)| changes the base semantics so that floating point literals have type \verb|clx.float|. This is useful because some OpenCL devices do not support double-precision floating point, or impose a significant performance penalty for its use. For applications where this higher precision is not necessary, this is convenient. Similarly, a base where integer literals are represented using true integers, rather than 32-bit integers, might be useful in  applications where avoiding accidental overflows is more important than performance. True integers can be implemented as an active type (not shown).

Indeed, to even use \verb|double| in OpenCL, an appropriate \verb|#pragma| must be inserted. The base detects uses of \verb|double| anywhere inside the function

\subsubsection{Composability and Interoperability}\label{safety}
Because a type can only exert control over typechecking and translation of operations where an expression of that type is the primary operand, extensions defined using AT\&T can not interfere with one another by construction. This does not imply that there are no \textbf{interoperability} issues to consider, however. A type may need to know about other types (e.g. pointer types need to know about integer types) and if this is done without consideration of future extensions, it may be difficult to integrate data produced by one type system with another. These issues cannot easily be addressed by a language design, however.

\section{Examples}\label{examples}

@@

\section{Discussion}\label{discussion}
Static type systems are powerful tools for programming language design and implementation. By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs. This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions. Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples). Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.

>>>>

As seen in Listing \ref{mapout}, we are implementing complex numbers internally using OpenCL vector types, like \verb|int2|. Let us look first at \verb|trans_New| on lines \ref{cplx}.54-\ref{cplx}.60, where new complex numbers are translated to vector literals by invoking \verb|VecLit| from the active target, which internally will generate the necessary OpenCL string for that form. The \verb|trans_type| method of the \verb|ace.ActiveType| is used to associated a type in the target language with an active type. For it to be possible to reason compositionally about the correctness of compilation, all complex numbers must translate to terms in the target language that have this target type \cite{tdc}. Ace supports a mode where this \emph{representational consistency} is dynamically checked during compilation (this requires that the active target know how to assign types to terms in the target language, which can be done only for our OpenCL target as of this writing).

The translation methods for attributes and binary operations proceed in a straightforward manner. The context provides a method, \verb|trans|, for recursively determining the translation of subexpressions as needed. Of note is that the translation methods can assume that typechecking succeeded. For example, the implementation of \verb|trans_Attribute| assumes that if \verb|node.attr| is not \verb|'ni'| then it must have been \verb|'i'| on line \ref{cplx}.14, consistent with the implementation of \verb|type_Attribute| above it. Typechecking and translation are separated into two methods to emphasize that typechecking is not target-dependent and to allow for typing rules that require generating types for hypothetical terms, where the overhead of the translation might be better avoided.

\subsection{Active Bases}\label{abases}
Each generic function is associated with an active base, which is an object implementing the \verb|ace.ActiveBase| interface. The active base specifies the \emph{base semantics} of that function. It controls the semantics of statements and expressions that do not have a clear ``primary'' subexpression for the dispatch protocol to defer to. A base is handed control over typechecking of statements and expressions in the same way as active types: via \verb|type_|$X$ and \verb|trans_|$X$ methods. It also initializes the context and governs the semantics of variables. Each function can have a distinct base semantics.

Literals are the most prominent form given their semantics by an active base. Our example active base, \verb|clx.base|, assigns integer literals the type \verb|clx.int32| while floating point literals have type \verb|clx.double|, consistent with the semantics of OpenCL and C99. The \verb|clx.base| object is an instance of \verb|clx.Base| that is pre-constructed for convenience. Alternative bases can be generated via different constructor arguments. For example, \verb|clx.Base(flit_t=clx.float)| is a base semantics where floating point literals have type \verb|clx.float|. This is useful because some OpenCL devices do not support double-precision numbers, or impose a significant performance penalty for their use. Indeed, to even use the \verb|double| type in OpenCL, a flag must be toggled by a \verb|#pragma| (The OpenCL target we have implemented inserts this automatically when needed, along with some other flags, and annotations like \verb|kernel|).  Similarly, for some applications, avoiding accidental overflows is more important than performance.  Arbitrary-precision integers can be implemented as an active type (not shown) and the base generates the constructor call form, described above, around integers to support this.

Another role that the active base can play is to provide operators that do not need to be imported. The semantics of \verb|get_global_id| and \verb|printf|, for example, are provided by \verb|clx.base|. In fact, the base provides a richer semantics for them than the underlying implementation, as we will show in the following section. It is not strictly necessary that the base provide these -- they could have been imported from \verb|clx| and used by their qualified name. A base which does not provide them without import can be generated by passing the \verb|primitives=False| flag in.

\subsection{Active Targets}\label{atargets}
An active target, which we show being used in the translation logic described above, is an instance of \verb|ace.ActiveTarget|. It has two primary roles in Ace: 1) it provides an API for code generation to a target language, providing term and type constructors for use during translation (and, to support representational consistency checks, a type system implementation); 2) it provides a bridge to the underlying interoperability API when a generic or concrete function is launched directly from Python, as described in Sec. \ref{implicit}.

An active type or active base can support multiple targets if desired by simply examining it during translation (e.g. by performing capability checks using Python's \verb|hasattr| function) and providing alternative code paths for different combinations of capabilities. Alternatively, an active backend can implement the interfaces available in a different back end and generate code for them in a different way. Our implementation of OpenCL's type system as a collection of active types and an active base does not explicitly support cross-compilation to a CUDA target (though it would be possible to modify it to do so), but our CUDA target does provide partial support for the OpenCL code generation API.

\subsection{Compositional Reasoning}\label{safety}
Every statement and expression in an Ace function is governed by exactly one active type, determined by the dispatch protocol, or by the single active base associated with the Ace function. The representational consistency check ensures that compilation is successful without requiring that types that abstract over other types (e.g. container types) know about their internal implementation details. Together, this permits compositional reasoning about the semantics of Ace functions even in the presence of many extensions. This stands on contrast to many prior approaches to extensibility, where extensions could insert themselves into the semantics in conflicting ways, making the order of imports matter (see Sec. \ref{related}).

\section{Expressiveness}\label{examples}
Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|. These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions. These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown. Some of the examples in the next section support this target in addition to the OpenCL target that we have been focusing on in this paper.

\subsection{Growing a Statically-Typed Python Inside an Ace}
Ace comes with a target, base and type implementing Python itself: \verb|ace.python.python|, \verb|ace.python.base| and \verb|ace.python.dyn|. These can be supplemented by additional active types and used as the foundation for writing actively-typed Python functions. These functions can either be compiled ahead-of-time to an untyped Python file for execution, or be immediately executed with just-in-time compilation, just like the OpenCL examples we have shown. Some of the examples in the next section support this target in addition to the OpenCL target that we have been focusing on in this paper.

@@

\section{Discussion}\label{discussion}
Python does not truly prevent extensions from interfering with one another because it lacks, e.g., data hiding mechanisms. One type could change the implementation of another by replacing it's methods, for example.

Static type systems are powerful tools for programming language design and implementation. By tracking the type of a value statically, a typechecker can verify the absence of many kinds of errors over all inputs. This simplifies and increases the performance of the run-time system, as errors need not be detected dynamically using tag checks and other kinds of assertions. Many parallel programming abstractions are defined in terms of, or benefit from, a type system that enforces a communication protocol, ensures the consistency of data and simplifies the dynamics of the run-time system (see Section \ref{usecases} for examples). Because \textbf{verifiability} and \textbf{performance} are key criteria and static typing is a core technique, Ace is fundamentally statically-typed.
